---
title: "Problem Set 2: A scaffolded multiple-page webscraping assignment"
output: 
  pdf_document:
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: sentence
---

```{r echo = FALSE, messages = FALSE, include=FALSE}
library(tidyverse)
theme_set( theme_minimal() )
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 5,
                      fig.height = 3,
                      out.width = "5in", 
                      out.height = "3in", fig.align = "center")

library( rvest )
library(purrr)
library(xml2)
```

## Our motivation

One of our group members attended a school district in Massachusetts, where the majority of students are White and the community leans toward a progressive agenda. They began to wonder whether the district’s stated values truly aligned with how they treat students of color. Consequently, our research question is:

RQ: Are black, hispanic, and multiracial students disciplined at disproportionately higher rates than their white peers?


To investigate this, we analyzed student discipline data. Because data collected during the pandemic was scant, we limited our scope to the last four years: 2022–2025.

We found that the school district remains predominantly white (72.5%), followed by asian students (8.8%), multiracial students (7.5%), hispanic students (7%), and black students (4%). Other races and ethnicities each represent less than 1% of the population, so we will focus our analysis on these most prevalent groups.


## Our goals

To answer the research question, we are scraping Student Discipline Data Reports from the Massachusetts Department of Elementary and Secondary Education (DESE) for three specific school districts: Concord K-8, Carlisle K-8, and Concord-Carlisle High School (CCHS), and the state wide data of Massachusetts.

Because these school districts are small and share a region, we are merging them into one group: the "Concord-Carlisle region." Since Concord K-8 and Carlisle K-8 are generally primary feeder schools for CCHS, this merger might provide a comprehensive view of the K-12 trajectory for students in this area. We will compare this aggregated data against the state-wide student discipline report, which encompasses K-12 enrollment data for all of Massachusetts.


\newpage

## Scraping: 

In this step, we scrape Student Discipline Data Reports for 2025 from four different units: Concord K-8, Carlisle K-8, Concord-Carlisle High School (CCHS), and Massachussets. The scope is extended into 4 years in the dedicated script: "PSet2_q3.R".
```{r messages = FALSE, results='hide'}

# We'll scrape 16 tables in total of 4 different units through the years 
# 2022-2025. Below we show a few examples of urls from 4 different units 
# (1 state + 3 school districts)

# commonwealth of massachusetts discipline totals for 2025
url_ma = paste0("https://profiles.doe.mass.edu/ssdr/default.aspx?orgcode=00000000&",
          "orgtypecode=0&=00000000&&fycode=2025")

# school district: Concordk8 discipline totals for 2025
url_conc = paste0("https://profiles.doe.mass.edu/ssdr/default.aspx?orgcode=00670000&",
            "orgtypecode=5&=00670000&&fycode=2025")


# school district: Carlislek8 discipline totals for 2025
url_carl = paste0("https://profiles.doe.mass.edu/ssdr/default.aspx?orgcode=00510000&",
            "orgtypecode=5&=00510000&&fycode=2025")

# school district: CCHS discipline totals for 2025
url_chs = paste0("https://profiles.doe.mass.edu/ssdr/default.aspx?orgcode=06400000&",
           "orgtypecode=5&=06400000&&fycode=2025")


#read the html from the page
discipline_ma = read_html(url_ma)
discipline_conc = read_html(url_conc)
discipline_carl = read_html(url_carl)
discipline_chs = read_html(url_chs)

#find the tables
discipline_table_ma = html_nodes( discipline_ma, "table" )
discipline_table_conc = html_nodes( discipline_conc, "table" )
discipline_table_carl = html_nodes( discipline_carl, "table" )
discipline_table_chs = html_nodes( discipline_chs, "table" )

#read the tables to identify which chunk we want
html_table(discipline_table_ma)
html_table(discipline_table_conc)
html_table(discipline_table_carl)
html_table(discipline_table_chs)

#extract the table we want
discipline_table_ma = html_table(discipline_table_ma[[4]], fill=TRUE)
discipline_table_conc = html_table(discipline_table_conc[[4]], fill=TRUE)
discipline_table_carl = html_table(discipline_table_carl[[4]], fill=TRUE)
discipline_table_chs = html_table(discipline_table_chs[[4]], fill=TRUE)

#now we have tables of disciplined students in 2025 from 4 different units
discipline_table_ma
discipline_table_conc
discipline_table_carl
discipline_table_chs

```

## Formatting:

In this step, we do some formatting and merge the pages scraped. We add a column with each unit name; we convert data type into character, so we're able to merge the for datasets; and remove columns with zero.
```{r messages = FALSE, results='hide'}

# before merging we have to add a column with the name of each unit
ma <- discipline_table_ma %>%
  mutate(unit_name = "Massachusetts")

conc <- discipline_table_conc %>%
  mutate(unit_name = "Concord")

carl <- discipline_table_carl %>%
  mutate(unit_name = "Carlisle")

cchs <- discipline_table_chs %>%
  mutate(unit_name = "CCHS")


# when trying to merge, we were getting this error: "! Can't combine `..
# 1$Students Disciplined` <character> and `..2$Students Disciplined` <integer>."
# so we converted everything into one class type before merging the tables
ma   <- map(ma, as.character)
conc <- map(conc, as.character)
carl <- map(carl, as.character)
cchs <- map(cchs, as.character)

# we used bind_rows to merge the tables, full_join doesn't merge more than 2 tables
merged_units <- bind_rows(ma, conc, carl, cchs)


# we removed the columns below because "Cells that contain 0.0% may be percentages
# that are more than 0.0% but less than 0.05%"so there was no data to be used
cols_to_remove <- c(
  "% Expulsion", 
  "% Alternate Setting", 
  "% Students with a School-Based Arrest", 
  "% Students with a Non-Arrest Law Enforcement Referral"
  )
merged_units <- merged_units %>% select(-all_of(cols_to_remove))

```


## Re-factor:

We now refactor the column containing demographic information about the students and edit column names to shorter version.
```{r messages = FALSE, results='hide'}

# refactor student group categories
merged_units <- merged_units %>%
  mutate(
      `Student Group` = fct_recode(
        `Student Group`,
        "AIAN"        = "American Indian or Alaska Native",
        "Black"       = "Black or African American",
        "Hispanic"    = "Hispanic or Latino",
        "Multiracial" = "Multi-Race, Not Hispanic or Latino",
        "NHPI"        = "Native Hawaiian or Other Pacific Islander",
        "High_Needs"  = "High Needs",
        "EL"          = "English Learners",
        "Low_SES"     = "Low Income",
        "SWD"         = "Students with Disabilities"
      )
    )


# make column names shorter
merged_units <- merged_units %>%
  rename(
    Group                 = "Student Group",
    Students              = "Students",
    Disciplined           = "Students Disciplined",
    ISS_pct               = "% In-School Suspension",
    OSS_pct               = "% Out-of-School Suspension",
    Emergency_removal_pct = "% Emergency Removal"
  )

```

\newpage

## Functions for raw html and refactoring

In this step, we save the raw HTML of each scraped unit into a local folder. Next, we define a function that parses these raw HTML files and converts them into structured tables. After validating this function on a single unit, we use map() to bulk-process all four units. Finally, we iterate a refactoring function across the four resulting tables to standardize column names and shorten values in demographics. 

Note: The script "PSet2_q4.R" performs a similar process, but it processes all 16 tables (4 units across 4 years).

```{r messages = FALSE, results='hide'}
# we need to create files to store the raw html of our pages
dir.create("raw_htmls", showWarnings = FALSE)

# raw HTML objects into a list
html_list <- list(
  "ma"   = discipline_ma,
  "conc" = discipline_conc,
  "carl" = discipline_carl,
  "cchs"  = discipline_chs
)

# file paths one for each unit, they are all from 2025
file_paths <- paste0("raw_htmls/unit_", names(html_list), ".xml")

# save raw htmls 
walk2(html_list, file_paths, write_html)

# function that takes raw html and returns full table
from_raw_html_to_table <- function(unit_name) {
  
  raw_html <- 
    read_file(
    file.path(paste0("raw_htmls/unit_", unit_name, ".xml")))
  discipline = read_html(raw_html)
  discipline_table = html_nodes( discipline, "table" )
  html_table(discipline_table)
  discipline_table = html_table(discipline_table[[4]], fill=TRUE)
  
  return(discipline_table)
  
}

# this returns the state wide table for Massachusetts from 2025
from_raw_html_to_table("ma")

# this returns the school district data table for Carlisle from 2025 
from_raw_html_to_table("carl")

# this returns the school district data table for CCHS from 2025
from_raw_html_to_table("cchs")

# this returns the school district data table for Concord from 2025 
from_raw_html_to_table("conc")

all_units = list("ma", "conc", "carl", "cchs")

# converting all raw html pages into tables for the 4 units
all_tables <- map(all_units, from_raw_html_to_table)

# function to refactor all tables
refactoring_table = function( table ) {

cols_to_remove <- c(
  "% Expulsion", 
  "% Alternate Setting", 
  "% Students with a School-Based Arrest", 
  "% Students with a Non-Arrest Law Enforcement Referral"
  )
table <- table %>% select(-all_of(cols_to_remove))

table <- table %>%
  mutate(
      `Student Group` = fct_recode(
        `Student Group`,
        "AIAN"        = "American Indian or Alaska Native",
        "Black"       = "Black or African American",
        "Hispanic"    = "Hispanic or Latino",
        "Multiracial" = "Multi-Race, Not Hispanic or Latino",
        "NHPI"        = "Native Hawaiian or Other Pacific Islander",
        "High_Needs"  = "High Needs",
        "EL"          = "English Learners",
        "Low_SES"  = "Low Income",
        "SWD"         = "Students with Disabilities"
      )
    )

# rename the variables
table <- table %>%
  rename(
         Group                 = "Student Group",                                  
         Students              = "Students",                                       
         Disciplined           = "Students Disciplined",                           
         ISS_pct               = "% In-School Suspension",                         
         OSS_pct               = "% Out-of-School Suspension",                    
         Emergency_removal_pct = "% Emergency Removal"                        
      )
}

all_tables <- map(all_tables, refactoring_table )

```

In this step, we were asked to test our existing function using a new page from the DESE website. However, because our function is designed to process raw HTML rather than URLs, we created a helper function that fetches a URL and provides the raw HTML required for testing on a page that has not yet been scraped.

```{r messages = FALSE, results='hide'}
# In order to test a different table from the website, one that is not already 
# stored as raw html, we need to create a function that takes url 

get_main_page = function(unit_id, orgtype, year) {
  url <- str_glue(
    "https://profiles.doe.mass.edu/ssdr/default.aspx?orgcode={unit_id}&",
    "orgtypecode={orgtype}&={unit_id}&&fycode={year}"
  )
  discipline       <- read_html(url)
  discipline_table <- html_nodes(discipline, "table")
  html_table(discipline_table[[4]])
}

# Test 1 with data from 2022 about school district Eastham (00850000)
discipline_eastham_2022 <- get_main_page(unit_id = "00850000", orgtype = 5, 
                                     year = 2022 )
# run refactoring function
discipline_eastham_2022 <- refactoring_table( discipline_eastham_2022 )

# Test 2 with data from 2024 about the school district Concord-Carlisle (06400000)
discipline_conc_2024 <- get_main_page( orgtype = 5, 
                                   unit_id = "06400000", year = 2024 )
discipline_conc_2024 <- refactoring_table( discipline_conc_2024 )
discipline_conc_2024

```

\newpage

## Concept: expand_grid  

Based on the Rdocumentation site, `expand_grid` creates a tibble from all combinations of inputs. If we were to just use `expand_grid`⁠ on year, school, and school_ID we would get every possible combination of those three inputs. This is not what we want because school and school_ID are matching pairs. Using left_join() is what preserves the pairing between school and school_ID.  


## Concept: what happens if we delete "data/"

What the "data/" part does is tell the script to look to a specific folder within the working directory called "data". By removing it, the script would save and look for the .xml files directly in the working directory that was set instead of within the "data" folder. 

## Concept: walk2

We use `walk2()` instead of `map()` because there are two inputs (pages$data and pages$file_name) plus a function. The "2" in `walk2()` indicates that it iterates over two inputs, whereas `map()` can only iterate over one.

Then, why don't we use `map2()` instead of walk2()? According to the 2023 tidyverse blog, purrr::walk() this way, the key difference between "walk2()" and "map2()" is the output:

`map()` and `map2()`: These return a list of results.

`walk()` and `walk2()`: These are used when we care about the side effect (such as writing a file) and do not return a list.

Since the goal of the script is to iterate over two inputs and write the HTML files, `walk2()` is the correct command to use.

References:
RDocumentation: https://www.rdocumentation.org/packages/tidyr/versions/1.3.2/topics/expand_grid
2023 tidyverse blog purr::walk() this way link: https://tidyverse.org/blog/2023/05/purrr-walk-this-way/


### The set of pages to scrape

Details on how and why we curated our list of units to be analyzed are described in section ## Problem Set Goals.


## Script that scrapes our data:

```{r, echo=FALSE, message=FALSE}

cat( readLines("PSet2_q3.R"), 
    sep = "\n" )
```


\newpage

## Script thats gets the 16 tables for our analysis:

```{r, echo=FALSE, message=FALSE}

cat( readLines("PSet2_q4.R"), 
    sep = "\n" )

```




\newpage

# Plotting the data on student discipline for Concord-Carlisle and Massachusetts

```{r, message=FALSE, warning=FALSE}
source("PSet2_q4.R")
# Before generating our plot, we needed to do some work on the data

# first: merge the three districts into one region called Concord-Carlisle and 
# limit the scope of race to Asian, Black, Multiracial, Hispanic, and White

plot_data <- tables_clean %>% 
  mutate(Region = ifelse(unit_name %in% c("Concordk8", "Carlislek8", "CCHS"), 
                         "Concord-Carlisle", "Massachusetts")) %>%
  group_by(year, Region, Group) %>%
  summarize(
    Disciplined = sum(Disciplined, na.rm = TRUE),
    Students    = sum(Students, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(discipline_pct = (Disciplined / Students) * 100) %>%
  filter(Group %in% c("Asian","Black", "Multiracial", "Hispanic",
                    "White"))
```

# Plot

```{r, message=FALSE, warning=FALSE}
# time region vs mass
plot_data %>%
  ggplot(aes(x = year, y = discipline_pct, color = Group, group = Group)) +
  geom_line(size = 0.5) +
  geom_point(size = 1) +
  facet_wrap(~Region) +
    labs(
    title = "Discipline Rates by Student Race/Ethnicity",
    x = "Years",
    y = "Percentage Disciplined (%)",
    color = "Race/Ethnicity"
  ) +
  theme_minimal() 

```

## Our thoughts:

Between 2022 and 2025, students of color—specifically Black and Hispanic students—were disciplined at disproportionate rates in both the Concord-Carlisle region and across Massachusetts. When compared to the state average, Concord-Carlisle generally maintains a lower overall percentage of disciplinary actions across all racial and ethnic groups.

In Concord-Carlisle, disciplinary reports for Black and Hispanic students saw a decline in 2024; however, 2025 data indicates a significant spike in reported discipline for Black students.

